### 2019-02-14 response/response-demography-1.tex

7. It is not clear why exactly 22, 23, 25 years are used a cut-off
points.

[Response:]
I have added a detailed discussion of recall error in the Appendices intended for online 
publication.
Furthermore, I have changed the cut-off to be 22 years of marriage or more for all surveys 
to simplify.
Even with the stricter cut-off there are still some evidence of recall error,
especially in region TK, where the sex ratio was substantially biased against females 
before sex selection became available.

8. I understand the empirical reason for not increasing the number of
variables. However, a key variable that is essential to consider is
region. This is important because regional influences on birth spacing
and sex selection are very strong as established in the literature. Sex
selection and abortion is especially strong in certain regions but not
in others. The regional differentials in fertility transition,
contraceptive use, spacing, sex ratios, abortions are clearly documented
in the literature. So it is vital to include this.

[Response:]
I agree completely that the regional differences in fertility, son preference, and use 
of sex selection are important.
To try to capture these differences, I have divided India into four broad regions 
depending on degree of son preference and initial fertility level as in 
\citep{retherford03b}, although expanded to include all states in the surveys.
In addition to the power problem, some education/region/spell/period combinations have too 
small sample sizes for the estimations to converge.
The lack of convergence is especially an issue for the middle-education group at the
higher spells.
I, therefore, first present the all-India results and then focus on regional differences
for only the most educated women, who are the main users of sex selection.

[Part of the reason is the very low fertility in the southern areas. See Guilmoto
and Rajan 2001 and IIPS 2006 and check latest NFHS report on, for example, Tamil Nadu.
TFR of less than 2 is likely.]

It is also worthwhile to note that even some of the southern states appear to have
seen an increase in son preference over time, or, rather, daughter aversion
\citep{DiamondSmith2008}.

### 2019-02-14 paper/birth_spacing_lit.tex
Add notes on spacing literature

To read:

In birth_spacing:

In collection.bib
\citep{Rafalitnanana2000} 
\citep{Rahman1993}
\citep{Jayachandran2017}
\citep{Jayachandran2017a}
\citep{Lambert2016}

In collection.bib but maybe not that relevant
\citep{Basu2010} 
\citep{dasgupta95} 
\citep{Kugler2017}

@@

>>>>

@@

The gender of the last-born child is signiﬁcant and its sign is
consistent with son-preference. If the last birth was a girl, the
expected birth interval is about 3\% shorter than if it was a boy.

There is a signiﬁcant humpshaped trend in birth-spacing, with a maximum
in 1980. That birth intervals got shorter in the last 20 years might be
because rising nutritional standards for mothers allow them to support
shorter intervals. Alternatively, the rise in women’s labour force
participation may have encouraged clustering of fertility.

birth intervals increase until the sixth child is born.

Other things equal, birth-order exhibits a non-monotonic pattern, with
the shortest birth intervals preceding the birth of the fourth child.

Maternal education decreases both mortality and fertility, but has no
effect on birth-spacing.

Fertility decline in India seems to have set in from 1981. Despite this,
birth intervals have got shorter since about then.

\citep{Zavier2000}

NFHS 92-93 Kerala

Many couples in India never use a reversible method to delay or space births, and in-
stead adopt sterilization as their first and only method.

higher educational attainment (of either partner) independently increases the likelihood 
that a couple will have used a method to delay or space births, as does middle 
socioeconomic status.

\citep{Maitra2008}

We ﬁnd that the estimates of birth spacing on child mortality are different when we do not account for fertility selection.

Our analysis is based on the National Family Health Survey (NFHS) 1992–1993 data from the Indian province of the Punjab and the Demographic and Health Survey (DHS) 1991–1992 data from the Pakistani province of the Punjab.

India: There is evidence of a statistically signiﬁcant effect of birth spacing on child survival: an increase in the duration between child i and child i + 1 signiﬁcantly reduces the hazard of mortality of child i (equivalently increases the survival chances of the child).
for the sample of Pakistani households. As with the Indian case, there is evidence of maternal depletion/sibling competition effect in the Pakistani sample: an increase in the duration between child i and child i + 1 signiﬁcantly reduces the hazard of mortality of child i.

early child death may also result in a reduction in the duration between successive children because parents want to replace children that have died.
Our results show that the child replacement effect is signiﬁcant in the Indian sample, but not in the Pakistani sample. In other words, increased child survival increases birth spacing in India but not in Pakistan.

\citep{Makepeace2008}

NFHS 1992-1993 West Bengal

Our central result is that child mortality decreases with increases in
either prior or posterior spacing (although the size of the fall is
different).

There is evidence that uncorrected estimates under-estimate the effects
of prior and posterior spacing on child mortality.

the spacing between consecutive children, i.e. the age composition of
siblings, could capture the intensity of competition between successive
siblings better,

\citep{Bhargava2003}

NFHS 1992-93: Uttar Pradesh

indicator variables constructed on the basis of the stated preferences for sons and daughters indicated that unwanted fertility contributes to excess mortality of higher order births, especially of girls.

\citep{Davanzo2008}

Matlab

Shorter intervals are associated with higher mortality. Interval effects are greater if the
interval began with a live birth than with another pregnancy outcome.

previous studies have generally considered the preceding interval between births* the interbirth interval*as their measure of spacing

the appropriate measure to use when considering the effect of competition or disease transmission from another young child in the family

sometimes there is a pregnancy that resulted in an outcome other than a live birth (NLB) between two live births, in which case the interbirth interval will include two (or more) interpregnancy intervals.

Some of the hypotheses about why reproductive spacing may affect infant and child health and survival are related to the interpregnancy interval

Furthermore, an intervening NLB may reflect something about the mother’s health that may affect the health of her children

Some, but very little, of the effect is explained by the fact that shorter interbirth and inter-outcome intervals are associated with shorter gestations of pregnancy.

Our results give some credence to the maternal depletion hypothesis.
We see that very short inter-outcome intervals are generally more detrimental when they follow a live birth or stillbirth than when they follow a preceding miscarriage or induced abortion. Because of their longer gestation, live births and stillbirths should be more depleting than miscarriages or induced abortions.
The breastfeeding that follows a live birth leads to further maternal depletion. Indeed we find that the effects of short inter-outcome intervals are greatest when the preceding outcome was a live birth.

We also find significant negative effects of subsequent short interpregnancy intervals on child survival (and have investigated this in a way that avoids reverse causality): a child is much more likely to die between ages 1 and 5 if the mother became pregnant again before this index child’s first birthday (RR02.31, pB0.001).

\citep{Haughton1995}

Vietnam Living Standards Survey (VLSS) 1992-1993

This article assesses the strength of son preference in Vietnam, as reflected infertility behavior. It
formulates and estimates a proportional hazards model applied to birth intervals, and a contraceptive
prevalence model, using household survey datafrom 2,636 ever-married women aged 15-49 with at
least one living child who were interviewedfor the Vietnam Living Standards Survey 1992-1993.

 Two tests are applied to the data in an attempt to discern the presence of son preference. The hazard
model test estimates the risk (hazard) of having another
child at any point in time; if the hazard is lower for families with a son (or sons),
the implication is that son preference is present. The contraceptive prevalence test
determines whether families with a son (or sons) are significantly more likely to use
contraceptives; if the answer is yer, a presumption may be made that son preference is present.

The central finding of this study is that son preference,
as reflected in household behavior, is strong in Vietnam.
This finding is surprising in the sense that Vietnamese
women are not secluded but are major participants in
the labor force and are relatively well educated.

\citep{Milazzo2018}

This paper is the ﬁrst to show that morbidity and mortality among adult women in India can be partially explained by son preference. First, I show suggestive evidence that women with a ﬁrst-born girl have lower survival. Second, consistent with prior literature, I ﬁnd that having a ﬁrst-born girl leads to fertility behaviors medically known to hurt women’s survival. Third, I show new evidence of eﬀects of a ﬁrst-born girl on a mother’s likelihood and severity of anemia. These outcomes are severely aggravated with each successive female birth

I pool the three available rounds of the India National Family Health Survey (NFHS) conducted in 1992/93, 1998/99, 2005/06, and consider the cross-sectional sample of women ages 15 to 39.

I show nonparametric evidence that the share of living women whose ﬁrst birth is female is a decreasing function of their age (or, alternatively, of the number of years since ﬁrst birth). This decreasing pattern starts around age 24 and continues to decrease below the biological range, stabilizing for women in their mid-30s

Although indirect, this evidence is suggestive of lower survival among women with ﬁrst girls due, at least partly, to son preference.

I estimate the eﬀects of having a ﬁrst-born girl on several fertility-related outcomes, including the probability of desiring more children, being sterilized, having undergone ultrasound, having had a terminated pregnancy, short birth spacing, and the number of children ever born.

I provide original evidence that women with ﬁrst-born girls are more likely to develop anemia. Speciﬁcally, a ﬁrst-born daughter increases the probability of being moderately or severely anemic by 5.4 percent. The additional eﬀect of a second and third-born girl (conditional on previous female births) is equivalent to 9 and 17.6 percent, respectively. Predetermined health or current pregnancy or breastfeeding status do not appear to drive these results.

\citep{Gangadharan2003}

We use an accelerated hazard model to estimate the duration between successive births and our results indicate that son preference exists only for the Indian community in South Africa.

Indian households are observed to have a higher duration between children following the birth of a son, irrespective of the number of children they already have. For the rest of the population, there is very little evidence of son preference.

For the Indian households, regardless of the number of existing sons, the duration is higher after a son. Additionally, an increase in the number of sons also significantly increases the duration between the second and the third children.

\citep{Kim2010}

Indonesia

among earlier cohorts, women who are more educated tend to have shorter birth intervals than those who are less educated and that the opposite is true among later cohorts.

The empirical results based on the ﬁrst hypothesis show that the level of industrialization of the local economy does not have a differential effect on birth spacing across educational groups. However, empirical results using measures of the availability of a family planning program show a differential effect on delaying births across educational groups.

women’s education had an impact on the second birth interval mainly through changing the availability of contraceptives rather than through change in the demand for children over the period 1974–90.

My ﬁndings are consistent with the hypothesis that more educated women are better at adopting new contraceptive technology.

\citep{Teachman1989}

USA

Our findings suggest
that a preference to balance the gender of children affects the timing of births, not a
preference for either sons or daughters. At parity 2, women with children of the same
sex time a third birth more rapidly than women with a boy and a girl. At parity 1,
women with a boy time second births more rapidly than women with a girl. This
seemingly anomalous finding is explained, however, by the fact that women with boys
are more likely than women with girls to be married at any point in time and thus
less likely to have disrupted fertility careers.

\citep{DiamondSmith2008}

Tamil Nadu, India

The southern Indian state of Tamil Nadu has experienced a dramatic decline in fertility, accompanied by a trend of increased son preference.

Findings suggest that daughter aversion, fuelled primarily by the perceived economic burden of daughters due to the proliferation of dowry, is playing a larger role in fertility decision-making than son preference. The desire for a son is often trumped by the worry over having many daughters. Women use various means of controlling the sex of their children, which in this study appear to be primarily female infanticide. It is important to distinguish between son preference and daughter aversion and to examine repercussions of low fertility within this setting.

In the last decade, Tamil Nadu – along with its neighbouring southern states – has achieved very low fertility. In the year 2005–2006, the Total Fertility Rate (TFR, the average number of children born to a woman over her lifetime) was estimated at 1.80, down from a TFR of 4.97 in 1974 (Dyson and Moore 1983, Guilmoto and Rajan 2001, International Institute for Population Sciences and ORC Macro 2006).

This impressive decline in fertility has been accompanied by worrisome evidence of the emergence of son preference in the past decade, which is reflected in increasingly disproportional sex ratios at birth and younger ages (Das Gupta and Bhat 1997, Basu 1999, Srinivasan 2005).

If strong daughter aversion is a prime driver of fertility limitation, it may put downward pressure on population growth, possibly helping to explain the recent trend toward very low fertility rates in Tamil Nadu. Although the phenomenon of daughter aversion has received less attention than son preference, there are a few past studies on it, primarily in Northern India, some which suggest that it may differ by religious group (Khanna 1997, Borooah and Iyer 2004). Our work here supports the idea that daughter avoidance, often described inaccurately as son preference, is increasing in rural Tamil Nadu and suggests that economic constraints, primarily due to increased use of dowry, are the cause.

\citep{Kishor1993}

India

Gender differentials in early childhood
mortality are best modeled by simultaneously
considering kinship structures and female labor force participation.

the analysis suggests that cultural factors may contribute more to explaining gender
differences in mortality across India than does female labor force participation.

this analysis unequivocally suggests that excess fe-
male mortality in India has both economic and
cultural roots and any explanation must con-
sider both economic and cultural factors.

\citep{Curtis1993}

Brazil

In this paper random-effects logistic models are used to analyze the effects of the
preceding birth interval on postneonatal mortality in Brazil, controlling for the
correlation of survival outcomes between siblings. The results are compared to those
obtained by using ordinary logistic regression. Family effects are found to be highly
significant in the random-effects model, but the substantive conclusions of the
ordinary logistic model are preserved. In particular, birth interval effects remain
highly significant

Birth interval effects are highly significant in Brazil, particularly in the less developed
Northeast region. This point is important because short birth intervals are relatively common
in Brazil, particularly in the Northeast.

\citep{Rahman1993}

We hypothesize that mothers in Matlab have a strong preference about the gender composition of their surviving children; in particular, they want several surviving sons and at least one surviving daughter.

the basic premise of this paper is that gender preference will have a greater influence on fertility where contraception is readily available and widely used; we expect to see little or no effect of gender preference on fertility in a population that does not use contraception.

In the earlier stages of family formation (i.e., in families with three or fewer surviving children), the number of surviving sons is a significant (negative) predictor of subsequent fertility. In the later stages (i.e., in families with four or more surviving children), the subsequent pace of reproduction is affected significantly not only by the number of sons but also by whether there is a son and by whether there is a daughter in the family. Our results show not only a very strong desire for sons, which is seen early in the family formation process, but also a desire for at least one daughter, which is seen at a more advanced stage.

The effect of gender composition of children is significantly stronger in the treatment area than in the comparison area. Moreover, whether the family has children of both genders seems to be more important in the treatment area than in the comparison area.

These differentials in the impact of gender preference on fertility as a result of the availability and use of contraception imply strongly that gender preference effects become more important when contraception is readily availableand when a population is more able to control its fertility, other things being equal.

\citep{Rafalitnanana2000}

Sub-Saharan Africa 20 countries

Examines preferred spacing as compared to observed spacing. In 5 of 20 countries the
actual spacing is significantly longer than the preferred spacing.

\citep{Lambert2016}

Exploiting original data from a Senegalese household survey, we provide evidence that fertility choices are partly driven by women's needs for widowhood insurance. We use a duration model of birth intervals to show that women most at risk in case of widowhood intensify their fertility, shortening birth spacing, until they get a son. Insurance through sons might entail substantial health costs since short birth spacing raises maternal and infant mortality rates.

\citep{Jayachandran2017}

This paper’s contribution is to directly estimate the causal relationship between family size and the desired sex ratio and to quantify an important cause of rising sex ratios. The challenge in estimating this effect is to isolate exogenous variation in the fertility level. The approach I use is to elicit sex composition preferences at different fertility levels: a hypothetical fertility level is specified to the survey respondent, and she is asked, given that fertility level, what is her preferred composition of boys and
girls. 5 By imposing the total number of children, one can characterize the respondents’ sex ratio preferences at different exogenously determined fertility levels.

\citep{Basu2010}

This article draws out some implications of son targeting fertility behavior and studies its determinants. We demonstrate that such behavior has two notable implications at the aggregate level: (a) girls have a larger number of siblings (sibling effect), and (b) girls are born at relatively earlier parities within families (birth-order effect). Empirically testing for these effects, we find that both are present in many countries in South Asia, Southeast Asia, and North Africa but are absent in the countries of sub-Saharan Africa. Using maximum likelihood estimation, we study the effect of covariates on son targeting fertility behavior in India, a country that displays significant sibling and birth-order effects. We find that income and geographic location of families significantly affect son targeting behavior.

\citep{Kugler2017}

Using data from nationally representative household surveys, we test whether Indian parents make trade-offs between the number of children and investments in education. To address the endogeneity due to the joint determination of quantity and quality of children, we instrument family size with the gender of the first child, which is plausibly random. Given a strong son preference in India, parents tend to have more children if the firstborn is a girl. Our instrumental variable results show that children from larger families have lower educational attainment and are less likely to be enrolled in school, with larger effects for rural, poorer, and low-caste families as well as for families with illiterate mothers.

\citep{Hotz1997}

This hazard approach provides a natural way to model incomplete histories, nonoccurrence of the event (a subsequent birth) and a natural set of covariates (current period values). It, however, does not solve several other problems. First, it provides
little insight into how to summarize the information in past and future covariates. Second, it does not solve the dynamic selection problem. To be in the sample of individuals on which we estimate the time between the first and second birth, one must have had a first birth. This is a selected sample. This simple hazard model provides no insight into the effects of that selection. Finally, this model has the unfortunate characteristic of mixing the parameters for the speed with which the event occurs with the parameters for whether or not the event occurs.
